---
title: "Food Preferences and Consumption Patterns"
author: "Vidya Sri L"
date: "2024-11-18"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

## Introduction

Dietary preferences and food consumption patterns are key aspects of
human behavior that have significant implications for public health,
nutrition, and the food industry. These preferences are shaped by a
complex interplay of demographic factors, health consciousness, cultural
influences, and individual lifestyle choices. Understanding these
patterns is crucial for designing effective public health interventions,
developing targeted marketing strategies, and addressing the diverse
needs of consumers.

This study aims to explore the intricate relationships between dietary
preferences and various influencing factors such as gender, health
consciousness, meal preferences, and eating habits. By employing
categorical data analysis techniques, the research seeks to uncover the
underlying associations that drive food choices. Specifically, the study
will analyze how gender affects dietary preferences, assess the impact
of health consciousness on food choices, examine the relationship
between meal preferences and eating out frequency, and develop
predictive models to forecast dietary preferences. Additionally,
advanced statistical methods, including multinomial logistic regression
and log-linear modeling, will be used to delve into more complex
relationships among multiple categorical variables related to food
consumption.

## Objectives

1.  Analyze Dietary Preferences: Examine the association between gender
    and dietary preferences, using odds ratio and relative risk to
    quantify differences.
2.  Assess Health Consciousness Impact: Determine how health
    consciousness influences food preferences through Chi-Square tests
    and visualizations.
3.  Evaluate Meal Preferences: Explore the link between meal preferences
    and frequency of eating out, applying likelihood ratio tests and
    heatmaps.
4.  Predict Dietary Preferences: Fit and evaluate logistic regression
    models to predict dietary preferences based on various predictors.
5.  Apply Multinomial Logistic Regression: Investigate shopping
    behaviors related to cooking frequency using multinomial logistic
    regression.
6.  Fit Log-Linear Model: Test the adequacy of the log-linear model in
    representing relationships among categorical variables related to
    food preferences.

## Formulation of Problem

The study aims to explore and analyze dietary preferences and food
consumption patterns using categorical data analysis. The focus is on
understanding how various factors such as gender, health consciousness,
meal preferences, and eating habits influence dietary choices among
respondents. The key issues include:

1.  Dietary Preferences by Gender: Assessing how gender affects
    preferences for vegetarian versus non-vegetarian diets.
2.  Health Consciousness and Food Choices: Investigating the
    relationship between health consciousness levels and preferred food
    types.
3.  Meal Preferences and Eating Out Frequency: Evaluating whether meal
    preferences are associated with how often individuals eat out.
4.  Dietary Preferences Prediction: Developing models to predict dietary
    preferences based on demo- graphic and behavioral factors.
5.  Multinomial and Log-Linear Modeling: Employing advanced statistical
    techniques to understand the complex relationships among multiple
    categorical variables related to food consumption.

## Questionnaire

The following questionnaire was designed to collect comprehensive data
on respondents’ food preferences, eating habits, and spending patterns.
It includes various types of questions to categorize responses and
assess key aspects of food consumption.

1.  What is your age group?

    • Under 18

    • 18-28

    • 29-38

    • 39-48

    • 49 and above

2.  What is your gender?

    • Male

    • Female

3.  What is your dietary preference?

    • Vegetarian

    • Non-Vegetarian

4.  What is your favorite type of cuisine?

    • Italian

    • Chinese

    • North Indian

    • South Indian

5.  How often do you eat out?

    • Daily

    • Weekly

    • Monthly

    • Occasionally

6.  Which meal of the day do you prefer the most?

    • Breakfast

    • Lunch

    • Dinner

    • Snacks

7.  How important is health to you when choosing food?

    • Important

    • Neutral

    • Does not care

8.  What is your preferred type of food?

    • Fast Food

    • Home-Cooked

    • Organic

    • Street Food

9.  How much do you typically spend on food per week?

    • Under 200

    • 200-500

    • 600-1000

    • 1000 and above

10. Where do you primarily shop for food?

    • Supermarket

    • Local Market

    • Organic Store

    • Online

11. How often do you cook meals at home?

    • Daily

    • Several times a week

    • Weekly

    • Occasionally

12. How often do you try new foods or cuisines?

    • Very Often

    • Often

    • Occasionally

    • Never

## Data Collection

### Description

Data was collected using an online survey tool, distributed among
participants from various age groups. The survey automatically recorded
respondents’ email addresses to ensure authenticity.

### Sample Size

A total of 131 respondents participated in the survey, providing diverse
insights into their food preferences and consumption habits.

### Methodology

The study utilizes a structured questionnaire to collect data on
respondents’ food preferences, eating habits, and spending patterns. The
survey was distributed online, targeting a diverse sample of
participants from various age groups, resulting in a total of 131
responses. The dataset was pre-processed to remove unnecessary columns,
and various statistical analyses were performed to meet the study’s
objectives.

• **Contingency Tables and Tests:** We created contingency tables to
explore the relationships between categorical variables, such as gender
and dietary preferences. The odds ratio and relative risk were
calculated to quantify differences, while Chi-Square tests were employed
to assess the significance of associations, such as between health
consciousness and preferred food types.

• **Visualization:** Heatmaps were generated to visually represent the
relationships between variables, providing a clearer understanding of
how different factors, such as health consciousness, influence food
preferences.

• **Logistic Regression:** A logistic regression model was fitted to
predict dietary preferences based on selected predictors, such as gender
and favorite cuisine. Model selection techniques, including forward
selection and backward elimination, were used to identify the most
important predictors.

• **Multinomial Logistic Regression:** This analysis was used to
investigate shopping behaviors related to cooking frequency, providing
insights into how often people shop at different places based on how
frequently they cook meals at home.

• **Log-Linear Modeling:** A log-linear model was fitted to test the
adequacy of representing relationships among categorical variables
related to food preferences, allowing us to understand the interactions
between multiple factors.

## Data Import and Initial Processing

```{r}
library(readxl)
#Load data
library(readxl)
data<- read_excel("/Users/apple/Downloads/CDA_Project (Responses).xlsx")
data <- data[,-c(1, 2)]
# Print first few rows of data
head(data)
```

## Analysis

```{r}
# Summary statistics for numerical variables
summary(data)

# Frequencies of categorical variables
table(data$age)
table(data$gender)
table(data$diet_pref)
table(data$fav_cuisine)
table(data$eat_out_freq)
table(data$meal_pref)
table(data$health_conscious)
table(data$pref_typefood)
table(data$spend_week)
table(data$shop_place)
table(data$cook_freq)
table(data$new_food)
```

**Demographics:** 69 females and 62 males. Majority are non-vegetarian
(94).

**Cuisine Preferences:** South Indian is the most popular cuisine (101).
Other favorites include North Indian (17), Italian (6), and Chinese (7).

**Meal Preferences:** Most preferred meals are dinner (51) and lunch
(49). Fewer respondents prefer breakfast (22) and snacks (9).

**Eating Out Frequency:** Weekly (63) is the most common frequency.
Other frequencies: Occasionally (26), Daily (27), and Monthly (15).

**Food Preferences:** Home-cooked food is highly preferred (92). Less
preference for fast food (16) and street food (11). Organic food has a
moderate preference (12).

**Shopping Habits:** Local Market (55) is the most popular shopping
place. Followed by Supermarket (45), Online (24), and Organic Store (7).

**Food Importance:** Most consider food important (73). A smaller number
are indifferent (10).

### Analysis of Gender and Dietary preferences:

```{r}
# Create contingency table with gender as the response variable
contingency_table_gen <- table(data$diet_pref, data$gender)

# Display the contingency table
print(contingency_table_gen)
```

The contingency table reveals the following dietary preferences: •
Females: 42 prefer non-vegetarian, and 27 prefer vegetarian. • Males: 52
prefer non-vegetarian, and 10 prefer vegetarian.

```{r}
# Define the values from the contingency table
a <- 42 # Females preferring non-vegetarian
b <- 27 # Females preferring vegetarian
c <- 52 # Males preferring non-vegetarian
d <- 10 # Males preferring vegetarian
# Calculate Odds Ratio
odds_ratio <- (a / b) / (c / d)
print(paste("Odds Ratio:", odds_ratio))
```

The odds ratio of approximately 0.299 indicates that the odds of females
preferring non-vegetarian food compared to vegetarian food are about 30%
of the odds for males. This suggests that females are significantly less
likely to prefer non-vegetarian food compared to males.

```{r}
# Calculate Relative Risk
risk_female_non_veg <- a / (a + b)
risk_male_non_veg <- c / (c + d)
relative_risk <- risk_female_non_veg / risk_male_non_veg
print(paste("Relative Risk:", relative_risk))
```

The relative risk of approximately 0.727 further supports this, showing
that females are about 73% as likely as males to prefer non-vegetarian
food.

### Analysis of Health consciousness and Preferred foodtype

```{r}
# Create a contingency table with health_conscious and pref_typefood
contingency_table_hf<- table(data$health_conscious, data$pref_typefood)

# Display the contingency table
print(contingency_table_hf)
```

```{r}
# Perform the Chi-Square test
chi_square_test <- chisq.test(contingency_table_hf)

# Display the results
print(chi_square_test)

# Critical value for alpha = 0.05 and df = 6
critical_value <- qchisq(0.95, df = 6)
print(critical_value)
```

The Chi-Square test results for the contingency table with
health_conscious and pref_typefood show a test statistic of 13.839 with
6 degrees of freedom, and a p-value of 0.03149. This p-value is below
the typical alpha level of 0.05, indicating that there is a
statistically significant association between health consciousness and
preferred food type.The critical value for the Chi-Square distribution
with 6 degrees of freedom at a 0.05 significance level is approximately
12.592. Since the test statistic 13.839 exceeds this critical value, we
reject the null hypothesis.

In other words, the differences observed in the frequency of preferred
food types across different levels of health consciousness are unlikely
to be due to chance, suggesting a meaningful relationship between health
consciousness and food preference.

#### Visualisation

```{r}
# Convert the contingency table to a data frame
contingency_df <- as.data.frame(contingency_table_hf)

# Load ggplot2 package
library(ggplot2)

# Create a heatmap with a color gradient from light to dark
ggplot(contingency_df, aes(x = Var2, y = Var1, fill = Freq)) +
geom_tile() +
scale_fill_gradient(low = "lightblue", high = "darkblue") +
labs(title = "Heatmap of Health Consciousness and Preferred Food Type",
x = "Preferred Food Type",
y = "Health Consciousness",
fill = "Frequency") +
theme_minimal()
```

The heatmap visually supports this finding by showing how the frequency
of each food preference varies with health consciousness, with some
categories (like “Important” in “Home-Cooked”) being more pronounced
than others.

### Analysis of Meal preference and eating outside frequently

```{r}
#Create contingency table
table = table(data$meal_pref, data$eat_out_freq)
table

#Likelihood Ratio Test
library(DescTools)

likelihood_ratio_test <- GTest(table)
print(likelihood_ratio_test)

# Calculate the degrees of freedom
df <- (nrow(table)- 1) * (ncol(table)- 1)
# Calculate the chi square tabulated value
p_value <- qchisq(0.95, df)
print(paste("Chi square tabulated value ", p_value))
```

Here, the likelihood ratio test (G-test), is used to determine whether
there is a significant association between meal preference and frequency
of eating out. The p-value is greater than the typical significance
level of 0.05, and Calculate G is less than table value, therefore we
fail to reject the null hypothesis which suggests that there is no
significant association between meal preference and frequency of eating
out at the 5% level.

#### Visualisation

```{r}
# Convert the contingency table to a data frame
table_df <- as.data.frame(table)
# Rename the columns
colnames(table_df) <- c("meal_pref", "eat_out_freq", "freq")
# Create a heatmap using ggplot2
ggplot(table_df, aes(x = eat_out_freq, y = meal_pref, fill = freq)) +
geom_tile() +
scale_fill_gradient(low = "white", high = "deeppink") +
labs(x = "Eat Out Frequency", y = "Meal Preference", fill = "Frequency") +
theme_classic()
```

The heatmap shows that “Lunch” is the most common meal for eating out,
especially on a weekly basis, as indicated by the darkest pink shade.
“Dinner” is also popular, particularly on an occasional and weekly
basis. Eating out daily is less common across all meal types. “Snacks”
and “Breakfast” have the lowest frequencies, with lighter shades across
all eating-out frequencies. Overall, weekly dining out, especially for
lunch, is the most frequent behavior.

### Logistic Regression and Model Evaluation on Dietary Preferences

#### Data Transformation and Initial Model Fitting

```{r}
data$diet_pref <- ifelse(data$diet_pref== "Vegetarian", 1, 0)
model=lm(data$diet_pref~.,data=data)

print(data)
```

The diet_pref column is converted into a binary variable where
“Vegetarian” is coded as 1 and others as 0. A linear model is fitted
with diet_pref as the response variable and all other variables as
predictors.

#### Test for Normality

**Shapiro-Wilk Test** • H0: The residuals are normally distributed • H1:
The residuals are not normally distributed.

```{r}
r=rstudent(model)
shapiro.test(r)
```

Here p-value 6.01e-05 \<0.05,we reject ho.Hence the residuals are not
normally distributed.

### Model Selection

#### Forward Selection

Forward selection helps in identifying the most important predictors for
diet_pref by incrementally adding variables that improve model
performance.

```{r}
fitstart=lm(data$diet_pref~1,data = data)
model <- lm(data$diet_pref~., data =data)
fwd=step(fitstart,direction = 'forward',scope = formula(model))
```

Gender and favourite cuisine are the most important predictors for
diet_pref.

### Backward Elimination

Backward elimination provides another method to refine the model by
removing predictors that do not contribute meaningfully to the
prediction of diet_pref.

```{r}
bwd=step(model,direction = "backward")
```

Gender and favourite cuisine are the most important predictors for
diet_pref selected through backward selection.

### Logistic Regression Model

```{r}
glm=glm(data$diet_pref~ gender + fav_cuisine,data=data,family = binomial)
summary(glm)
```

A generalized linear model (GLM) is fitted using gender and fav_cuisine
as predictors, with diet_pref as the binary outcome.The logistic
regression analysis shows that gender is a significant predictor of
dietary preference. Specifically, males are significantly less likely to
be vegetarian compared to females (Estimate = -1.20226, p = 0.00757).
The decrease from 155.95 (null deviance) to 138.44 (residual deviance)
suggests that the model with gender and fav_cuisine improves the fit
compared to a model with no predictors

### Model Evaluation

```{r}
# Load necessary libraries
library(caret) # For confusionMatrix and metrics
library(pROC) # For ROC curve and AUC

# Predict probabilities
pred_probs <- predict(glm, data, type = "response")
# Convert probabilities to binary predictions
pred <- ifelse(pred_probs > 0.5, 1, 0)
# Calculate confusion matrix
conf_matrix <- confusionMatrix(as.factor(pred), as.factor(data$diet_pref))
# Extract accuracy, sensitivity, specificity
accuracy <- conf_matrix$overall['Accuracy']
sensitivity <- conf_matrix$byClass['Sensitivity']
specificity <- conf_matrix$byClass['Specificity']
# Print accuracy, sensitivity, and specificity
print(paste("Accuracy:", round(accuracy, 4)))
print(paste("Sensitivity:", round(sensitivity, 4)))
print(paste("Specificity:", round(specificity, 4)))

# Calculate F1 Score
precision <- conf_matrix$byClass['Precision']
recall <- sensitivity
f1_score <- 2 * (precision * recall) / (precision + recall)
print(paste("F1 Score:", round(f1_score, 4)))

# Print AUC value
roc_curve <- roc(data$diet_pref, pred_probs)

auc_value <- auc(roc_curve)
print(paste("AUC:", round(auc_value, 4)))
```

The model shows good accuracy (77.1%) and is excellent at identifying
vegetarians (sensitivity: 95.74%). However, it struggles to correctly
identify non-vegetarians (specificity: 29.73%). The F1 Score (0.8571)
reflects a good balance between precision and recall, while the AUC
(0.6998) indicates moderate overall discrimination ability.

### ROC curve

```{r}
# Plot ROC curve
roc_curve <- roc(data$diet_pref, pred_probs)

auc_value <- auc(roc_curve)
plot(roc_curve, main = "ROC Curve")
abline(a = 0, b = 1, lty = 2, col = "red") # Add a diagonal line
```

The ROC curve shows that the model has moderate performance. The curve
is close to the diagonal, indicating that the model’s ability to
distinguish between classes is not strong. This suggests the model is
only somewhat better than random guessing, with room for improvement.

The model is very effective at detecting vegetarians but struggles with
accurately identifying non-vegetarians. The high sensitivity and F1
score suggest it’s well-tuned for situations where identifying
vegetarians is more critical, but the low specificity indicates it may
produce many false positives for vegetarians.

#### Fitting of Multinomial Logistic Regression:

```{r}
table1 = table(data$shop_place, data$cook_freq)
table1

# Load the necessary libraries
library(nnet)
# Convert the shop_place and cook_freq variables to factors
data$shop_place <- factor(data$shop_place)
data$cook_freq <- factor(data$cook_freq)

# Perform multinomial logistic regression analysis
multinom_shop_place <- multinom(shop_place~ cook_freq, data = data)

# Summarize the results
summary(multinom_shop_place)

# Extract the coefficients for each category
coef_shop_place <- coef(multinom_shop_place)
# Calculate the odds ratios for each category
or_shop_place <- exp(coef_shop_place)
print("Odds Ratios for shop_place:")

print(or_shop_place)
```

The multinomial logistic regression shows that people who cook more
frequently (especially weekly) are much more likely to shop at
supermarkets and organic stores. Those who cook less frequently, like
occasionally or several times a week, are more inclined to shop online.
The odds ratios indicate a strong preference for supermarkets and
organic stores among regular cooks, with the highest likelihood seen in
weekly cooks.

### Fitting of Log-Linear Model:

Null Hypothesis (H0): The log-linear model accurately represents the
relationships among the categorical variables in the data. In other
words, the model fits the data well and there are no significant
deviations from what is expected.

Alternative Hypothesis (H1): The log-linear model does not accurately
represent the relationships among the categorical variables.

```{r}
library(MASS)
# Create a contingency table and set dimension names
contingency_table <- table(data$age, data$pref_typefood, data$spend_week, data$new_food)
dimnames(contingency_table) <- list(
age = levels(data$age),
pref_typefood = levels(data$pref_typefood),
spend_week = levels(data$spend_week),
new_food = levels(data$new_food)
)
# Fit the log-linear model using loglm
loglin_model <- loglm(~ age + pref_typefood + spend_week + new_food, data = contingency_table)
loglin_model
```

The log-linear model fit well according to the statistics. The
Likelihood Ratio Chi-Square value is 138.39 with a p-value close to
1(\>0.05), indicating a good model fit. The Pearson Chi-Square value is
274.76 with a p-value of 0.079 (\>0.05), suggesting no significant
deviation between observed and expected frequencies. Overall, these
results imply that the log-linear model adequately represents the
relationships among the categorical variables of the data.

## Conclusion:

The study successfully explored dietary preferences and food consumption
patterns using categorical data analysis. Key findings include:

-   **Gender and Dietary Preferences:** Females were found to be
    significantly less likely to prefer non- vegetarian food compared to
    males, with both odds ratio and relative risk analyses supporting
    this conclusion.

-   **Health Consciousness and Food Choices:** A significant association
    was found between health consciousness and preferred food type,
    indicating that individuals who value health are more likely to
    prefer home-cooked or organic foods.

-   **Meal Preferences and Eating Out Frequency:** The likelihood ratio
    test suggested no significant association between meal preference
    and the frequency of eating out, indicating that these factors may
    be independent of each other.

-   **Predictive Modeling:** Logistic regression analysis identified
    gender and favorite cuisine as significant predictors of dietary
    preferences. The model demonstrated good accuracy in predicting
    vegetarians but struggled with specificity in identifying
    non-vegetarians.

-   **Multinomial and Log-Linear Modeling:** The multinomial logistic
    regression revealed strong preferences for supermarkets and organic
    stores among frequent cooks, while the log-linear model adequately
    represented the relationships among various categorical variables
    related to food consumption.

    Overall, the findings provide valuable insights into the factors
    influencing dietary preferences and food consumption patterns,
    offering potential applications in public health, marketing, and
    consumer behavior research. The study’s methodology and statistical
    analyses contribute to a deeper understanding of the complex
    interactions between demographic, behavioral, and psychological
    factors in shaping food choices.\

## Further Recommendation

Based on the findings of this study, several recommendations can be made
to better address the diverse dietary preferences and food consumption
patterns observed:

-   **Targeted Health Promotion Campaigns:** Public health initiatives
    should focus on increasing awareness about healthy food choices,
    especially among groups that show lower health consciousness.
    Tailored campaigns could encourage healthier eating habits,
    particularly for individuals who prefer fast food or street food.

<!-- -->

-   **Supporting Sustainable Eating Practices:** Considering the growing
    interest in health-conscious and organic food, policies that support
    sustainable farming practices and make organic products more
    accessible and affordable could align with consumer preferences and
    promote overall well-being.

-   **Developing Mobile Apps for Dietary Guidance:** Technology
    companies could develop or enhance mobile apps that provide
    personalized dietary advice based on user input regarding their food
    preferences, health consciousness, and eating habits. This could
    help users make informed food choices in real-time.
